Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-01 paper/paper.tex
finish thompson sketch

\newcommand{\isTy}{\textrm{ type}}
\newcommand{\isCtx}{\textrm{ ctx}}
\newcommand{\isSmall}{\textrm{ small}}
\newcommand{\isLinTy}{\textrm{ lin. type}}
\newcommand{\isLinCtx}{\textrm{ lin. ctx.}}

@@

Just as in \cref{fig:fingram}, we can show that particular strings match a Kleene
star. In \cref{fig:kleenestarderivation}, we demonstrate that \stringquote{ab}
matches $(a^{*} \otimes b) \oplus c$ through repeated application of the Kleene
star constructors. In this proof tree we expand some syntactic details that are often
left implicit. For instance, every linear context is parameterized by a nonlinear
context, which is syntactically separated by a semicolon. If the nonlinear
context is empty, we may often omit it; however, in this figure we make this
distinction explicit. Further consider the leaves of the tree. First, the
leaves involving the constructors are nonlinear terms, and thus they occur in
the nonlinear empty context. Then we explicit cast from
an nonlinear $\uparrow$-wrapped type into a linear one. Such casts may often be
made implicitly, such as in $\cref{fig:kleeneabstractproof}$.

@@

\paragraph{Equivalence}
In \cite{chom1963}, Chomsky differentiates between the weak generative capacity
of a formal grammar --- the set of strings accepted by it --- and its strong
generative capacity --- the set of ``structural descriptions'', or parse trees,
that it generates. Two grammars are said to be \emph{weakly equivalent} by
Chomsky if they
agree in their weak generative capacity, and likewise \emph{strongly equivalent}
if they agree in their strong generative capacity. We aim to recover each of
these notions of equivalence internally.

@@

\paragraph{Regular Expressions} The theory of regular expressions emerges as the
subtheory of \theoryabbv generated by
$\otimes$, $\oplus$, and Kleene star over the types $\bot$, $I$, and characters drawn from $\Sigma$.

@@

\subsection{The Dyck Grammar}
%% 3. Dyck language
\newcommand{\balanced}{\mathsf{balanced}}
\newcommand{\Dyck}{\mathsf{Dyck}}
Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
grammars (CFGs). In many ways, the smallest nontrivial CFG is the Dyck grammar
$\Dyck$ --- the grammar of balanced parentheses. Moreover, the Chomsky-Sch\"utzenberger
representation theorem \cite{CHOMSKY1959118} demonstrates that $\Dyck$, and grammars like it,
cut to the essence of what it means to be context free.

$\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$
and is defined in \cref{fig:dyckinductive}. The $\nil$ constructor shows that
the empty string is balanced, and the $\balanced$ constrcutors builds a balanced
parse by wrapping an already balanced parse in an additional set of parentheses
then following it with another balanced parse.

\begin{figure}
\begin{align*}
\data &~\Dyck : L~\where\\
      & \nil : \ltonl {\Dyck} \\
      & \balanced : \ltonl {(\stringquote{(} \lto \Dyck \lto \stringquote{)} \lto \Dyck \lto \Dyck)}
\end{align*}
\caption{Kleene Star as an inductive type}
\label{fig:dyckinductive}
\end{figure}

\steven{somewhere in here, sneak in the benchmark where the Dyck parser could
  handle 10's of thousands of characters in about 30s}

We construct a parser for $\Dyck$ by building a determinstic automaton $M$ such
that $\Trace_{M}(M.\init) \cong \Dyck$.

\Cref{fig:DyckAutomaton} presents a sketch of the automaton $M$. The traditional
approach to parse grammars like $\Dyck$ is to employ a stack to count how many
open parentheses have been seen so far. We may think of
$M$ as performing the same task as a stack machine, except instead of storing
the number of open parentheses in a distinguished stack of memory, $M$ operates
by storing this data in the current state. This is made feasible by specifying
$M$ almost identically to how we defined DFAs, except we omit the
finiteness condition on the states of the automaton.

\steven{Mention that $\Trace_{M}$ is iso to string. Mention the same for DFAs above }

% Unlike the finite automata seen thus far, $M$ has infinitely many states --- one
% for each $n : \mathbb{N}$ with an additional fail state. This diverges a bit
% from classical language theory, where the recognizer for $D$ would be encoded as a
% pushdown automaton with finitely many
% states and stack of memory. Such PDAs are encodable in our theory, although they
% are a little cumbersome to use. Moreover, they
% seem overfit to classical reasoning as they don't preserve the grammar up to
% strong equivalence, only weak equivalence in general. \steven{cite Lambek
%   automata and state that is is a secondary goal to preserve as much structure as possible}
% By moving the stack of memory (the nats) into the
% state, we obtain infinitely many states, which is fine bc the states need only
% be a type for us to interface with it.

\steven{TODO sketch proof of equivalence. Mention that it is mechanized}

\begin{figure}
  \begin{tikzpicture}[node distance = 17mm ]
    \node[state, initial, accepting] (0) {$0$};
    \node[state, right of=0] (1) {$1$};
    \node[state, right of=1] (2) {$2$};
    \node[right of=2] (3) {$\dots$};
    \node[state, below of=0] (fail) {$\mathsf{fail}$};

    \path[->] (0) edge[above, bend left] node{\stringquote{(}} (1)
              (1) edge[below, bend left] node{\stringquote{)}} (0)
              (1) edge[above, bend left] node{\stringquote{(}} (2)
              (2) edge[below, bend left] node{\stringquote{)}} (1)
              (2) edge[above, bend left] node{\stringquote{(}} (3)
              (3) edge[below, bend left] node{\stringquote{)}} (2)
              (fail) edge[loop right] node{\stringquote{(}, \stringquote{)}} (fail)
              (0) edge[left] node{\stringquote{)}} (fail);
  \end{tikzpicture}
  \caption{Automaton $M$ for the Dyck grammar}
  \label{fig:DyckAutomaton}
\end{figure}

% \subsection{LL(1)}
% %% 4. LL(1) parser

% \steven{TODO impute LL(1) example, mention that its weak equivalence is mechanized}

@@

\newcommand{\Turing}{\mathsf{Turing}}
\newcommand{\snoc}{\mathsf{snoc}}

\begin{figure}
\begin{align*}
\data &~\Turing~(t : \mathbb{Z} \to \Gamma): L~\where\\
      & \nil : \ltonl {\left(\LinPiTy {t} {\mathbb{N} \to \Gamma} {\Turing~t} \right)} \\
      & \snoc : \ltonl {\left( \LinPiTy {c}{\Sigma}{\LinPiTy {t} {\mathbb{N} \to \Gamma} {\Turing(\mathsf{push}(c , t)) \lto c \lto \Turing(t)}} \right)}
\end{align*}
\caption{Kleene Star as an inductive non-linear type}
\label{fig:turinginductive}
\end{figure}

We nonlinearly encode the tape of a (one-sided) Turing machine as
function from $\mathbb{N} \to \Gamma$, where $\Gamma$ is the tape alphabet.

We may then define a term of type
$\mathsf{parseTape} : \ltonl {\left( \stringg \lto \LinSigTy {t} {\mathbb{N} \to \Gamma} {\Turing(t)} \right)}$.
In this manner, all of the interesting computations that the Turing machine
would do occur solely inside of the nonlinear index. That is,
$\mathsf{parseTape}$ does not discern which strings are accepted by the Turing
machine, nor does it do any computation. $\mathsf{parseTape}$ simply loads the
input string into the initial spaces on the tape.

Once the string has been loaded onto the tape, we may further define the
following term to run the Turing machine.
\[
  \mathsf{run} : \ltonl {\left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \Turing(t) \lto \bigwith\limits_{\mathsf{fuel} : \mathbb{N}} \mathsf{Maybe} \left( \bigoplus\limits_{t : \mathbb{N} \to \Gamma} \bigoplus\limits_{b : \Bool} \bigoplus\limits_{\mathsf{trace} : \mathsf{TuringTrace}(b , t)} \Turing(t) \right) \right)}
\]

where $\mathsf{TuringTrace}$ is a nonlinear type denoting terminating traces
through the Turing machine. Note that $\mathsf{run}$ returns a $\mathsf{Maybe}$
type in a fuel parameter. Turing machines may not terminate. If they do
terminate within $\mathsf{fuel}$-many steps, then we return the trace along with
the current tape and a boolean denoting if the trace is accepting or rejcting.
If the trace does not terminate within the alloted amount of fuel, then we
timeout and return nothing.

@@

\section{Denotational Semantics}
\label{sec:semantics-and-metatheory}

To justify our intuitive explanation of linear types as formal
grammars, we will now define a \emph{denotational semantics} that
makes this mathematically precise. This semantics is the basis for our
prototype implementation in Agda.

\subsection{What is a formal grammar?}

\begin{definition}
  A \emph{formal language} $L$ is a function from strings to propositions.
  A \emph{formal grammar} $A$ is a function from strings to small sets.
\end{definition}

\max{todo: need to decide if we need this section here or if it's in the intro already}.

@@

\begin{definition}

\end{definition}

To handle our universe, we assume a universe of small sets in our
metatheory that is closed under all of our type formers.
\begin{definition}[Grammar Semantics]
  We define the following interpretations by mutual recursion on the
  judgments of \theoryabbv:
  \begin{enumerate}
  \item For each non-linear context $\Gamma \isCtx$, we define a set $\sem \Gamma$.
  \item For each non-linear type $\Gamma \vdash X \isTy$, and element
    $\gamma \in \sem\Gamma$, we define a set $\sem X \gamma$.
  \item For each linear type $\Gamma A \isLinTy$ and element $\gamma
    \in \sem\Gamma$, we define a formal grammar $\sem{A}\gamma$. We
    similarly define a formal grammar $\sem\Delta\gamma$ for each
    linear context $\Gamma \Delta\isLinCtx$.
  \item For each non-linear term $\Gamma \vdash M : X$ and $\gamma \in \sem{\Gamma}$, we define an element $\sem{M}\gamma \in \sem{X}\gamma$.
  \item For each linear term $\Gamma; \Delta \vdash e : A$ and $\gamma \in \sem{\Gamma}$ we define a parse transformer from $\sem{\Delta}\gamma$ to $\sem{A}\gamma$.
  \end{enumerate}
  And we verify the following conditions:
  \begin{enumerate}
  \item If $\Gamma \vdash X \isSmall$, then $\sem X \gamma$ is a small set.
  \item If $\Gamma \vdash X \cong X'$ then for every $\gamma$, $\sem{X}\gamma = \sem{X'}\gamma$
  \item If $\Gamma \vdash A \cong A'$ then for every $\gamma$, $\sem{A}\gamma = \sem{A'}\gamma$
  \item If $\Gamma \vdash M \cong M' : X$ then for every $\gamma$, $\sem{M}\gamma = \sem{M'}\gamma$
  \item If $\Gamma;\Delta \vdash e \cong e' : A$ then for every $\gamma$, $\sem{e}\gamma = \sem{e'}\gamma$
  \end{enumerate}
\end{definition}

For reasons of brevity, we present only a selection of cases here. The full details are in the appendix.\max{todo: put the full details in the appendix}.

Most of the non-linear semantics is standard \pedro{We should either cite something or add the non-linear semantics in the appendix}, so let us investigate in more
depth the linear semantics.
\begin{enumerate}
  \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
  \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
        $\sem X : \sem \Gamma \to \Set_{i}$
  \item A non-linear term $\Gamma \vdash e : X$ denotes a dependently typed function
        $\sem e : \prod_{\gamma : \sem \Gamma}\sem{X} \gamma$ \pedro{we have not defined what sections are}
  \item Linear contexts $\Gamma \vdash \Delta$ and types
        $\Gamma \vdash A : L_{i}$ both denote families of grammars
        $\sem \Gamma \to \Gr_{i}$
  \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
        transformers
        $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
\end{enumerate}

>>>>

@@

Just as in \cref{fig:fingram}, we can show that particular strings match a Kleene
star. In \cref{fig:kleenestarderivation}, we demonstrate that \stringquote{ab}
matches $(a^{*} \otimes b) \oplus c$ through repeated application of the Kleene
star constructors. In this proof tree we expand some syntactic details that are often
left implicit. For instance, every linear context is parameterized by a nonlinear
context, which is syntactically separate by a semicolon. If the nonlinear
context is empty, we may often omit it; however, in this figure we make this
distinction explicit. Further consider the leaves of the tree. First, the
leaves involving the constructors are nonlinear terms, and thus they occur in
the nonlinear empty context. Then we explicit cast from
an nonlinear $uparrow$-wrapped type into a linear one. Such casts may often be
made implicitly, such as in $\cref{fig:kleeneabstractproof}$.

@@

\paragraph{Equivalence}
In \cite{chom1963}, Chomsky differentiates between the weak generative capacity
of a formal grammar --- the set of strings accepted by it --- and its strong
generative capacity --- the set of ``structural descriptions'', or parse trees,
that it generates. Two grammars are said to be \emph{weakly equivalent} if they
agree in their weak generative capacity, and likewise \emph{strongly equivalent}
if they agree in their strong generative capacity. We aim to recover each of
these notions of equivalence internally.

@@

\paragraph{Regular Expressions} The theory of regular expressions emerges as a
subtheory of \theoryabbv by restricting the typing connectives to only use
$\otimes$ and $\oplus$ generated over $\bot$, $I$, and characters drawn from $\Sigma$.

@@

\subsection{LL(1)}
%% 4. LL(1) parser

\steven{TODO impute LL(1) example, mention that its weak equivalence is mechanized}

@@

\begin{figure}
\begin{align*}
\data &~\mathsf{Turing}~(t : \mathbb{Z} \to \Gamma): L~\where\\
      & \mathsf{nil} : \ltonl {\left(\LinPiTy {t} {\mathbb{Z} \to \Gamma} {\mathsf{Turing}~t} \right)} \\
      & \mathsf{snoc} : \ltonl {(\mathsf{Turing} \lto A^{*} \lto A^{*})}
\end{align*}
\caption{Kleene Star as an inductive non-linear type}
\label{fig:turinginductive}
\end{figure}

However, unlike
the traces through the automata we have seen until now, the encoding of Turing
machines

We nonlinearly encode the tape of the Turing machine as
function from $\mathbb{Z} \to \Gamma$, where $\Gamma$ is your tape alphabet. The
grammar of traces for a Turing machine linear consumes the input string and
writes it to the tape. From here, we nonlinear reason about machine transitions.
We execute the machine and track its current state as part of the nonlinear
indexing data.

%% Make sure to say which results are mechanized

@@

\section{Semantics and Metatheory}
\label{sec:semantics-and-metatheory}

\subsection{Formal Grammars as Linear Types}
\label{sec:formaltype}
\max{this should be in its own section on semantics}

@@

Most of the non-linear semantics is standard \pedro{We should either cite something or add the non-linear semantics in the appendix}, so let us investigate in more
depth the linear semantics.

\begin{enumerate}
  \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
  \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
        $\sem X : \sem \Gamma \to \Set_{i}$
  \item A non-linear term $\Gamma \vdash e : X$ denotes a section
        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$ \pedro{we have not defined what sections are}
  \item Linear contexts $\Gamma \vdash \Delta$ and types
        $\Gamma \vdash A : L_{i}$ both denote families of grammars
        $\sem \Gamma \to \Gr_{i}$
  \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
        transformers
        $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
\end{enumerate}
